Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-07 doc/mc-paper.tex
Updated

\newcommand{\simplify}[2]{#1|_{#2}}
\newcommand{\prov}{\mathit{Expand}}

@@

\begin{abstract}
Knowledge compilers convert Boolean formulas into a form for which
useful tasks, including model counting and weighted model counting,
can easily be performed.  Existing knowledge compilers, however,
provide no guarantee that the generated representation accurately
describes the original formula.

@@

We have developed a program that can generate POG representations and
equivalence proofs from the d-DNNF graphs generated by a leading
knowledge compiler.  It scales to all but the largest formulas from a
recent model counting competition.  We also present two proof
checkers---a fast checker that can handle the large proofs generated,
and a formally verified checker written in the \lean{} proof system.
\end{abstract}

@@

@@

For each variable $x \in \varset$,
  we define {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
  negation of $x$. An
  assignment $\assign$ can be viewed as a set of literals, where
  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
$\obar{\lit} = x$ when $\lit = \obar{x}$.

  We highlight the following special classes of Boolean formulas:
  \begin{description}
    \item[Normal form:] A formula where negation is applied only to variables.
    \item[Conjunctive Normal Form (CNF):] A normal form formula where  disjunction is applied only to literals.
    \item[Partitioned-Operation:] A formula satisfying the following for all conjunction and disjunction operations: 
      \begin{itemize}
      \item The arguments to a conjunction must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to a disjunction must have disjoint models.  That is, operation
        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{itemize}
  \end{description}
     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned-operation formula.
     A formula in conjunctive normal form can be represented by a set
     of {\em clauses}, each of which is a set of literals.  Each
     clause then represents the disjunction of the literals, and the formula is then the conjunction of its clauses.

  For Boolean formula $\phi$, the {\em ring evaluation problem} (REP) is to compute:
  \begin{eqnarray}
    \rep(\phi) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
\label{def:labeling}
\end{definition}

A wide variety of important tasks for Boolean functions can be
expressed as ring evaluation problems.  The 
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$
It can be cast as an REP by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.  
Ring evaluation of formula $\phi$ will then compute the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ will then
yield the number of models.

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow 
assigning weights $W$ to both the variables and their negations, such
that there can be variables $x$ with $W(x) + W(\obar{x}) \not = 1$.  We can convert this more general form into a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and then computing the final result as $\rep(\phi)\;
\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a randomized test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1) = \rep(\phi_2)$
if they are logically equivalent.  If they are not equivalent, then
the probability that $\rep(\phi_1) \not = \rep(\phi_2)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Therefore, ring evaluation can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

Possible generalizations of this framework are discussed in Section~\ref{sect:future}.

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following:
\begin{description}
\item[Negation] $\rep(\boolnot \phi) = \mulident - \rep(\phi)$
\item[Partitioned Conjunction] $\rep\left(\Pand_{i=1,k} \phi_i\right) = \prod_{i=1,k} \rep(\phi_i)$
\item[Partitioned Disjunction] $\rep\left(\Por_{i=1,k} \phi_i\right) = \sum_{i=1,k} \rep(\phi_i)$
\end{description}
\end{proposition}

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  When $(u,v) \in E$,
node $v$ is said to be a {\em child} of node $u$.
The in- and out-degrees of node $u$ are defined as $\indegree(u) = | E \cap (N \times \{u\}) |$, and
$\outdegree(u) = | E \cap (\{u\} \times N) |$.
Node $u$ is said to be {\em terminal} if $\outdegree(u) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $u$ is said to be {\em nonterminal} if $\outdegree(u) > 0$.  A nonterminal node is labeled by one of three Boolean operations:
$\boolnot$, $\pand$, or $\por$.  Node $u$ can be labeled with $\boolnot$ only if $\outdegree(u) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
There is a unique {\em root node} $r$ such that $\indegree(r) = 0$.

As can be seen, a POG is simply a way to represent a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
$\phi(u)$ as the formula denoted by node $u$ and write $\phi(P) = \phi(r)$ for root $r$ of $P$.

We define the {\em size} of POG $P$, written $|P|$ to be the sum of
the number of nonterminal nodes and the number of edges.  Ring
evaluation of $\phi(P)$ can be performed with at most $|P|$ ring
operations by traversing the graph from the terminal nodes up to
the root, computing a value $\rep(u)$ for each node $u$.

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  disjunction node $v$ in a d-DNNF can have at most two children $u_1$ and $u_0$ for which there must be a {\em splitting variable} $x$ such that
  any total assignment $\assign$ satisfying $\phi(u_1)$ has $\assign(x)=1$, while any total assignment satisfying $\phi(u_0)$ has
  $\assign(x)=0$.  Our generalization is provided to allow an
  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

As is shown in Figure~\ref{fig:chain}
two programs are involved in producing a certified compilation of a formula:
\begin{itemize}
\item The {\em proof generator} generates a CRAT file that both describes a POG and provides a proof that this POG is equivalent to the input formula.
  This compiler might start with the output of another knowledge compiler.
\item The {\em proof checker} ensures that all of the proof conditions are satisfied.  
\end{itemize}

The CRAT format draws its inspiration from the LRAT format for Boolean
formulas~\cite{lrat} and the QRAT format for quantified Boolean
formulas (QBF)~\cite{heule:JAR2014}.  The following are its key properties:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Each such declaration implicitly adds an extension variable representing the POG node and a set of {\em defining} clauses
  encoding the conjunction or disjunction operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring
  $k\,(k-1)/2$ proofs of disjointness for larger values of $k$.
\end{itemize}

@@

Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
The checker is provided with the input formula as a separate file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $\lit$
is represented by a signed integer, with $-v$ being the complement of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that clause $C$ can only reference clauses $C'$ such that $C' < C$.

The first set of proof rules are similar to those in other clausal
proofs.
Clauses can be added via RUP addition
(command {\tt a}), with a sequence of antecedent clauses (the
``hint'').
Similarly for clause deletion (command {\tt dc}).

The declaration of a $\pand$ operation has the form:
\begin{center}
\begin{tabular}{ccccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\tt 0} \\
\end{tabular}
\end{center}
where $i$ is a new clause ID, $v$ is a positive integer that does not
correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$  
integers representing literals of existing variables.  
This declaration implicitly causes $k+1$ clauses to be added to the proof:
\begin{center}
\begin{tabular}{cccccc}
\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
  $i$ & $v$ & $-\lit_1$ & $-\lit_2$ & $\cdots$ & $-\lit_k$\\
  $i\!+\!1$ & $-v$ & $\lit_1$  \\
  $i\!+\!2$ & $-v$ & $\lit_2$  \\
  & $\ldots$ \\
  $i\!+\!k$ & $-v$ & $\lit_k$  \\
\end{tabular}
\end{center}
The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A $\pand$ operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then becomes a way
of representing constant $0$.

The declaration of a $\por$ operation has the form:
\begin{center}
\begin{tabular}{ccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} 
\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
\end{tabular}
\end{center}
where $i$ is a new clause ID, $v$ is a positive integer that does
not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
signed integers representing literals of existing variables.  Hint $H$
consists of a
sequence of clause IDs.
This declaration implicitly causes three clauses to be added to the proof:
\begin{center}
\begin{tabular}{cccc}
\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
  $i$ & $-v$ & $\lit_1$ & $\lit_2$ \\
  $i\!+\!1$ & $v$ & $-\lit_1$ \\
  $i\!+\!2$ & $v$ & $-\lit_2$ \\
\end{tabular}
\end{center}
The hints must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$.

@@

Observe that the defining clauses for the conjunction and disjunction
operations provide a full Tseitin encoding of the operations.  Letting
$\exvarset$ denote the set of added extension variables,  any
total assignment $\assign\colon \varset \rightarrow \{0,1\}$ has a unique extension,
which we also call $\assign$, with $\assign\colon (\varset \cup \exvarset) \rightarrow \{0,1\}$.

@@

At the completion of the proof, the following conditions must hold:
\begin{itemize}
\item All of the input clauses have been deleted.
\item There should be exactly one clause left that was added via RUP
  addition, and this should be a unit clause consisting of the root literal $r$.
\end{itemize}
The sequence of clause addition steps provide a proof that any
total assignment $\assign$ to the input variables that satisfies all of the
clauses must, when extended, assign $\assign(r) = 1$.  Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must, when extended,
assign $\assign(r) = 0$.

The sequence of operator declarations, asserted clauses, and deleted
clauses represents a systematic transformation of the input formula
into a POG\@.  Validating each of these steps serves to prove that the
POG is logically equivalent to the input formula.

@@

Figure \ref{fig:eg4:all} illustrates an example formula and shows how
the CRAT file declares its POG representation.  The input formula (A)
consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
$x_4$.  The generated POG (B) has six nonterminal nodes representing four
conjunctions and two disjunctions.  We name these by the node type (product or sum) and the extension variable.
The first part of the CRAT file
(C) declares these nodes using clause IDs that increment by three or
four, depending on whether the node has two children or three.
The last two nonzero values in the sum declarations are hints providing the required mutual exclusion proof.

We step through a detailed example to provide a better understanding of the CRAT proof framework.
Figure
\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
defined by the POG operation declarations.  These do not appear in the
CRAT file.  Referring back to the declarations of the sum nodes in
Figure \ref{fig:eg4:all}, we can see that the declaration of node
$s_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
\ref{fig:eg4:all}(A) that these two clauses resolve to the clause
$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $s_7$
have disjoint models.  Similarly, node $s_{10}$ is declared as having
clause IDs 16 and 19 as hints.  These resolve to the clause
$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
$s_{10}$ have disjoint models.

Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $s_{10}$ is implied by the input clauses, i.e.,
any total assignment $\assign$
satisfying the input clauses must extend to one with $\assign(s_{10}) = 1$.
Working backward, we can see that
clause 29 indicates that node $p_8$ will be implied by the input
clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
hints for clause 36.

>>>>

\newcommand{\makenode}[1]{\mathbf{#1}}
\newcommand{\nodeu}{\makenode{u}}
\newcommand{\nodev}{\makenode{v}}
\newcommand{\nodes}{\makenode{s}}
\newcommand{\nodep}{\makenode{p}}
\newcommand{\noder}{\makenode{r}}

\newcommand{\simplify}[2]{#1|_{#2}}
\newcommand{\prov}{\mathit{Prov}}

@@

\begin{abstract}
  Knowledge compilers convert Boolean formulas into a form 
  that enables efficient evaluation of properties of the formula,
  including
  weighted and unweighted model counting.
 Existing knowledge compilers, however,
provide no guarantee that the generated representation accurately
describes the original formula.

@@

We have developed a program that can generate POG representations and
equivalence proofs from the d-DNNF graphs generated by a leading
knowledge compiler.  We also present two proof
checkers---a fast checker that can handle the large proofs generated,
and a formally verified checker written in the \lean{} proof system.
Our generation and checking tool chain scales to all but the largest formulas from a
recent model counting competition.  
\end{abstract}

@@

For each variable $x \in \varset$,
  we define the {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
  negation of $x$. An
  assignment $\assign$ can be viewed as a set of literals, where
  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
$\obar{\lit} = x$ when $\lit = \obar{x}$.

  We highlight the following special classes of Boolean formulas:
  \begin{description}
    \item[Normal form:] A formula where negation is applied only to variables.
    \item[Conjunctive Normal Form (CNF):] A normal form formula where  disjunction is applied only to literals.
    \item[Partitioned-Operation:] A formula satisfying the following for all conjunction and disjunction operations: 
      \begin{itemize}
      \item The arguments to a conjunction must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to a disjunction must have disjoint models.  That is, operation
        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{itemize}
  \end{description}
     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned-operation formula.
     A formula in conjunctive normal form can be represented by a set
     of {\em clauses}, each of which is a set of literals.  Each
     clause represents the disjunction of the literals, and the formula is the conjunction of its clauses.

  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} (REP) is to compute:
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, the summation $\sum$ is computed using operation $\radd$, and the product $\prod$ is computed using operation $\rmul$.
\label{def:labeling}
\end{definition}

A wide variety of important tasks for Boolean functions can be
expressed as ring evaluation problems.  The 
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a REP by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.  
Ring evaluation of formula $\phi$ will then compute the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ will then
yield the number of models.

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow 
indepdently assigning weights $W(x)$ and $W(\obar{x})$ to each variable $x$, potentially having
$W(x) + W(\obar{x}) \not = 1$.
We can convert this more general form into a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and then computing the final result as $\rep(\phi, w)\;
\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a randomized test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.  If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Therefore, ring evaluation can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

Generalizations of this framework are discussed in Section~\ref{sect:future}.

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{description}
\item[Negation] $\rep(\boolnot \phi, w) = \mulident - \rep(\phi, w)$
\item[Partitioned Conjunction] $\rep\left(\Pand_{i=1,k} \phi_i, w \right) = \prod_{i=1,k} \rep(\phi_i, w)$
\item[Partitioned Disjunction] $\rep\left(\Por_{i=1,k} \phi_i, w\right) = \sum_{i=1,k} \rep(\phi_i, w)$
\end{description}
\end{proposition}

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.  A nonterminal node is labeled by one of three Boolean operations:
$\boolnot$, $\pand$, or $\por$.  Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.

As can be seen, a POG is simply a way to represent a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
$\phi(\nodeu)$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi(\nodeu))$.

We define the {\em size} of POG $P$, written $|P|$ to be the sum of
the number of nonterminal nodes and the number of edges.  Ring
evaluation of $P$ can be performed with at most $|P|$ ring
operations by traversing the graph from the terminal nodes up to
the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
The final result is then $\rep(\noder, w)$.

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  disjunction node $\nodev$ in a d-DNNF can have at most two children $\nodeu_1$ and $\nodeu_0$ for which there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(u_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
Our generalization is provided to allow an
  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

@@

As is shown in Figure~\ref{fig:chain}
two programs are involved in producing a certified compilation of a formula:
\begin{itemize}
\item The {\em proof generator} produces a CRAT file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
\item The {\em proof checker} ensures that all of the proof conditions are satisfied.  
\end{itemize}

The CRAT format draws its inspiration from the LRAT format for Boolean
formulas~\cite{lrat} and the QRAT format for quantified Boolean
formulas (QBF)~\cite{heule:JAR2014}.  The following are its key properties:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Each such declaration implicitly adds an extension variable $u$ representing each  POG node $\nodeu$ and a set of {\em defining} clauses
  encoding the conjunction or disjunction operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring
  $k\,(k-1)/2$ proofs of disjointness for larger values of $k$.
\end{itemize}

@@

Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
The checker is provided with the input formula as a separate file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $\lit$
is represented by a signed integer, with $-v$ being the logical negation of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that any clause identifier $C'$ given in the hints when adding clause $C$ must have $C' < C$.

The first set of proof rules are similar to those in other clausal
proofs.
Clauses can be added via RUP addition
(command {\tt a}), with a sequence of antecedent clauses (the
``hints'').
Similarly for clause deletion (command {\tt dc}).

The declaration of a $\pand$ operation has the form:
\begin{center}
\begin{tabular}{ccccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\tt 0} \\
\end{tabular}
\end{center}
Symbol `{\tt p}' refers to ``product.''
Integer $i$ is a new clause ID, $v$ is a positive integer that does not
correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$  
integers representing literals of existing variables.  
This declaration implicitly causes $k+1$ clauses to be added to the proof, defining extension variable $v$ to be the conjunction of its arguments:
\begin{center}
\begin{tabular}{cccccc}
\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
  $i$ & $v$ & $-\lit_1$ & $-\lit_2$ & $\cdots$ & $-\lit_k$\\
  $i\!+\!1$ & $-v$ & $\lit_1$  \\
  $i\!+\!2$ & $-v$ & $\lit_2$  \\
  & $\ldots$ \\
  $i\!+\!k$ & $-v$ & $\lit_k$  \\
\end{tabular}
\end{center}
The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A $\pand$ operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then becomes a way
of representing constant $0$.

The declaration of a $\por$ operation has the form:
\begin{center}
\begin{tabular}{ccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} 
\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
\end{tabular}
\end{center}
Symbol `{\tt s}' refers to ``sum.''
Integer $i$ is a new clause ID, $v$ is a positive integer that does
not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
signed integers representing literals of existing variables.  Hint $H$
consists of a
sequence of clause IDs.
This declaration implicitly causes three clauses to be added to the proof, defining extension variable $v$ to be the disjunction of its arguments:
\begin{center}
\begin{tabular}{cccc}
\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
  $i$ & $-v$ & $\lit_1$ & $\lit_2$ \\
  $i\!+\!1$ & $v$ & $-\lit_1$ \\
  $i\!+\!2$ & $v$ & $-\lit_2$ \\
\end{tabular}
\end{center}
The hints must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$.

@@

The sequence of operator declarations, asserted clause, and
clause deletions represents a systematic transformation of the input formula
into a POG\@.  Validating each of these steps serves to prove that the
POG is logically equivalent to the input formula.

Observe that the defining clauses for the conjunction and disjunction
operations provide a full Tseitin encoding of the operations.  For any node $\nodeu$ and associated extension variable $u$,
we can therefore assume that any total assignment $\assign$ to the input variables will also assign
a value to the extension variables such that $\assign(u) = 1$ if and
only if $\assign \in \modelset(\nodeu)$.

@@

At the completion of the proof, the following conditions must hold:
\begin{itemize}
\item All of the input clauses have been deleted.
\item There should be exactly one clause left that was added via RUP
  addition, and this should be a unit clause consisting of the root literal $r$.
\end{itemize}
The sequence of clause addition steps provide a proof that
$\modelset(\phi) \subseteq \modelset(\noder)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$, and therefore 
$\modelset(\noder) \subseteq \modelset(\phi)$.

@@

Figure \ref{fig:eg4:all} illustrates an example formula and shows how
the CRAT file declares its POG representation.  The input formula (A)
consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
$x_4$.  The generated POG (B) has six nonterminal nodes representing
four conjunctions and two disjunctions.  We name these by the node
type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
extension variable.  We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
  The first part of the CRAT file (C) declares
these nodes using clause IDs that increment by three or four,
depending on whether the node has two children or three.  The last two
nonzero values in the sum declarations are hints providing the
required mutual exclusion proof.

We step through a detailed example to provide a better understanding of the CRAT proof framework.
Figure
\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
defined by the POG operation declarations.  These do not appear in the
CRAT file.  Referring back to the declarations of the sum nodes in
Figure \ref{fig:eg4:all}, we can see that the declaration of node
$\nodes_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
\ref{fig:eg4:all}(A) that these two clauses resolve to the clause
$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
clause IDs 16 and 19 as hints.  These resolve to the clause
$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
$\nodes_{10}$ have disjoint models.

Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $\nodes_{10}$ is implied by the input clauses, i.e.,
any total assignment $\assign$
satisfying the input clauses must have $\assign(s_{10}) = 1$.
Working backward, we can see that
clause 29 indicates that variable $p_8$ will be implied by the input
clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
hints for clause 36.

Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
delete the input clauses.  Consider the first of these, deleting
input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
The proof proceeds by first assuming that the clause is false, requiring
$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hints then consist of unit
clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
clauses that cause unit propagation.  Hint clauses 8 and 10 force the
assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 forces
$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
completing the RUP proof.  The deletion hints for the other input
clauses follow similar patterns---they work from the bottom nodes of
the POG upward, showing that any total assignment that falsifies the clause
must assign $\assign(s_{10}) = 0$.

Deleting the asserted clauses is so simple that we do not show it.  It
involves simply deleting the clauses from clause number 35 down to
clause number 25, with each deletion using the same hints as were used
to add that clause.  In the end, therefore, only the defining clauses
for the POG nodes and the unit clause asserting $s_{10}$ remain,
completing a proof that the POG is logically equivalent to the input
formula.
